Typed Clojure

### 2013-07-25 schemeworkshop13.tex
preprint
Add dealing with null

\documentclass{sigplanconf}

@@

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

@@

Our basic approach is to make our array types \emph{bivariant}. Array types
look like \lstinline|(Array2 w r)| and
are reminicent of functions or pipes: having a contravariant parameter for input (writing)
and a covariant parameter for output (reading).

@@

The type inferred for the local \lstinline|n| is (Array2 Nothing Number)
which tells the type system: it is never safe to write to this array, but
it is safe to assume \lstinline|Number|s can be read from this array.

@@

>>>>

\documentclass[preprint]{sigplanconf}

@@

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Typed Clojure}
%\subtitle{Subtitle Text, if any}

@@

Our basic approach is to make our array types \emph{bivariant}. Array types
look like \lstinline|(Array2 w r)| and
are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
and a covariant parameter for output (reading).

@@

The type inferred for the local \lstinline|n| is \lstinline|(Array2 Nothing Number)|
which tells the type system: it is never safe to write to this array, but
it is safe to assume \lstinline|Number|s can be read from this array.

@@

\subsection{Handling null}

Probably the most common pain point in Java programming is dealing with \emph{null}.
It is crucial that Typed Clojure deals with \emph{null} intelligently so that
all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
We also want to represent \emph{null} in a way that is natural to Clojure programmers
and approximates their mental model when reasoning about Clojure code.

Firstly, we separate the concepts of \emph{null} and reference types at the type level.
This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
We provide a singleton type \lstinline|nil|, which contains just the value \lstinline|nil|,
Clojure's equivalent of \emph{null}.

Armed with general unions and the type \lstinline|nil|, 
we can make Typed Clojure's reference types \emph{non-nullable}.
A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
in Typed Clojure.

Typed Racket already provides what is needed to write such types; indeed this is a repurposing
of a common Typed Racket idiom of approximating an Option or Maybe type by
making a union that includes \emph{\#f}, Racket's only false value.
However, like handling covariant arrays, the subtlety lies in our interactions with foreign
Java code.

A set of conversion rules define how to convert a Java type signature into a null-safe
Clojure version. These rules are designed to be conservative and can be explicitly overridden
by programmers. Method parameters are non-nullable, while return types are nullable.
Field types are nullable. Constructors parameters are non-nullable, and return types
are also non-nullable: it is a Java invariant that all constructors return an instance
of the Class they are constructing.
